---
title: "AQI vs Mobility Project: Data Preparation"
author: "Anjelica Martinez"
date: "05/08/2020"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(easypackages)
libraries("sjmisc", "tidyverse", "lubridate", "readr", "plyr")
```

## Introduction

The historic COVID-19 pandemic has caused nearly 50% of the world population to be subject to government “stay-at-home” and self-isolation orders. This sudden, drastic reduction in the mobility of the population has generated a natural experiment that tests the effect a near halt of mobility for individuals and drastic reduction of commercial operations would have on air quality. The metric used for this study is the mean daily mobility of individuals in a given US county.
  
Based on this criteria, we selected five major counties that had the highest level of mobility reduction as well as five other counties that had fewer reduction in mobility. A major county is defined as being in the top 40 most populous counties in the US, per the US Census Bureau. 

The time period of this study is February 28 through March 27, the time during which some US counties enacted “shelter-in-place” and the self-isolation orders to control the community spread of COVID-19. 

The five counties with the greatest reduction in citizen mobility issued “shelter-in-place” orders during this period. 

The five counties with the least mobility reduction had no “shelter-in-place” orders during the studied time period and can serve as a control group. Our goal is to determine whether decrease in mobility correlates with increase in air quality and how strong of a relationship is there for these 10 counties.

**Counties of Interest**

Counties that decreased mobility **98-100%** between Feb. 28 and Mar. 27, 2020. [New York Times](https://www.nytimes.com/interactive/2020/04/02/us/coronavirus-social-distancing.html).

* Arlington, VA
* District of Columbia, DC
* New York, NY
* San Francisco, CA
* Santa Clara, CA

Most populous counties that decreased mobility **30% or less** between between Feb. 28 and Mar. 27, 2020 [New York Times](https://www.nytimes.com/interactive/2020/04/02/us/coronavirus-social-distancing.html).

* Miami-Dade, FL
* Dallas, TX
* Harris, TX
* Maricopa, AZ
* San Bernardino County, CA

## Data Sources
### Air Quality Data
We collected our air quality data from the [EPA's free API service](https://www.epa.gov/outdoor-air-quality-data/air-quality-index-daily-values-report). From the website, "The AirData Air Quality Index Summary Report displays an annual summary of Air Quality Index (AQI) values for counties or core based statistical areas (CBSA). Air Quality Index is an indicator of overall air quality, because it takes into account all of the criteria air pollutants measured within a geographic area. Although AQI includes all available pollutant measurements, you should be aware that many areas have monitoring stations for some, but not all, of the pollutants." You can read the documentation [here](https://www.epa.gov/outdoor-air-quality-data/about-air-data-reports).

### Mobility Data
We used the mobility statistics data collected by 
[Descarte Labs ](https://github.com/descarteslabs/DL-COVID-19). The data represents "the distance a typical member of a given population moves in a day, at the US admin1 (state) and admin2 (county) level)". Follow the link to the github repository to read their documentation.

### Time Range Adjustment
Because the mobility dataset only has data from March 1st we decided to omit air quality data from the months of January and February. Thus, our analyses only covers from March 1st to April 18th.

## Data Preparation
### Mobility Data Prep
First I cleaned and tidied up the *mobility* index csv file from Decartes Labs.

**Variables of Interest:**

* `m50`: The median of the max-distance mobility for all samples in the specified region
* `m50_index`: The percent of normal m50 in the region, with normal m50 defined during 2020-02-17 to 2020-03-07

If you would like to read more about the Descarte Labs mobility index, you can find the documentation [here](https://github.com/descarteslabs/DL-COVID-19).

```{r message=FALSE, warning=FALSE}
# Import Messy Data
df <- read.csv("../Scripts/data/messy/us_mobility.csv", stringsAsFactors = FALSE)
(df <- as_tibble(df))
```

I renamed the columns to make it easier to understand what values they represented. I changed `admin1` to `state` and `admin2` to `county`. I then changed the date column from character format to data format.
```{r}
# Rename columns
df <- dplyr::rename(df, state = admin1, county = admin2)

# Change character dates to R date format
df$date <- as.Date(df$date)

head(df)
```

I created three dataframes for easier analysis:
* `decrease`: mobility index values for counties that decreased mobility *98-100%* between Feb. 28 to Mar. 27
* `increase`: mobility index values for  counties that decreased mobility *30% or less* between Feb. 28 to Mar. 27
* `df`: mobility index values for *all* counties

#### Decrease Dataframe
```{r}
# Create dataframe that contains counties that decreased mobility 98-100% between Feb. 28 to Mar. 27
decrease <- df[(df$county %in% c("New York County", "Arlington County", "San Francisco County", "Santa Clara County")), ]
```
The mobility index had a peculiar entry. For Washington, D.C. they had the state listed as Washington D.C and the county column was left blank. I changed the state value to District of Columbia and added the District of Columbia value to all the District of Columbia state entries.
```{r}
dc_data<- df[(df$state == "Washington, D.C."), ]
decrease <- bind_rows(decrease, dc_data)

# Replace Washington, DC with District of Columbia
decrease$state[decrease$state == "Washington, D.C."] <- "District of Columbia"

# Add District of Columbia value
decrease[197:245, 5] <- "District of Columbia"

head(decrease)
```
Because the AQI dataframe excludes the county suffix, I removed the 'county' suffix from all the country entries for cohesiveness and easier analysis.
```{r}
# Remove County from county name
decrease$county <- gsub(" County", "", decrease$county)

# Check that District of Columbia values have been added
decrease[(decrease$county == "District of Columbia"), ]

# Check decrease dataframe
head(decrease)
```
#### Increase Dataframe
This dataframe contains the mobility index values for counties that decreased mobility *30% or less* between Feb. 28 to Mar. 27. 

```{r}
# Create dataframe that contains Counties that decreased mobility 30% or less between Feb. 28 to Mar. 27
increase <- df[(df$county %in% c("Miami-Dade County", "Dallas County", "Harris County", "Maricopa County", "San Bernardino County")), ]
```
This dataset contained several states that shared the same counties names. Here, I removed the states that we were not analyzing as well as the county suffix.
```{r}
increase <- filter(increase, !(state %in% c("Alabama", "Arkansas", "Iowa", "Missouri", "Georgia")))

# Remove "county" from county name
increase$county <- gsub(" County","", increase$county)
```

I then combined these two datasets together to form `sample_df` that only contained the data for the 10 counties of interest. I kept the original intact dataframe as `df` in case we decided to include more counties.
```{r}
# Create df with counties of interest (our sample)
sample_df <- bind_rows(decrease, increase)
```
I decided to rename the variables when I saved the .Rdata file in order to make it easier identification in future analyses scripts.
```{r}
# Rename with clearer variables
all_mobility_data <- df
decreased_mobility <- decrease
increased_mobility <- increase
all_selected_counties <- sample_df

# Check variables
head(all_mobility_data)
head(decreased_mobility)
head(increased_mobility)
head(all_selected_counties)

# Save as Rdata file
# save(decreased_mobility, increased_mobility, all_mobility_data, all_selected_counties, file = "mobility.RData")

# Check that data loads
# load("mobility.RData")
```

### AQI Data Prep

**Variables of Interest:**

* `AQI`: average air quality index for that day
* `Pollutant names (e.g. Ozone, PM25)`: each have their average daily value
* `main_pollutant`: from the original data, represented which pollutant was the main pollutant affecting AQI

The API system provided by EPA required us to download each counties daily average aqi and pollutant values individually. I used an apply function (loop) to quickly process all the csv files and combine them into one dataset called `df`.
```{r message=FALSE, warning=FALSE}
# Apply loop to import csv files at once
files <- list.files(path = "../Scripts/data/messy/daily_aqi_values", pattern = "*.csv", full.names = T)
df <- ldply(files, read_csv)

# Change to data format
df$Date <- as.Date(df$Date)

# Rearrange columns
df <- move_columns(df, County, .after = "Main Pollutant")

#Change to tibble format
df <- as_tibble(df)

head(df)
```
Next I adjusted the columns to tidy format (lowercase, minimal) and renamed `Overall AQI` to `aqi`.
```{r}
# Rename Columns
names(df)
(names(df) <- c("date", "aqi", "main_pollutant",
               "county", "site_name", "site_id",
               "source", "ozone", "pm25", "co",
               "so2", "pm10", "no2"
               ))
```

Before saving the .Rdata file I renamed the df variable to aqi_df so that when it was imported it would not replace another variable called df.
```{r}
# rename variable
aqi_df <- df

# Save Workspace
# save(aqi_df, file = "daily_aqi_and_pollutants.RData")
```

### Merge AQI and Mobility Dataframes
```{r}
# Load .Rdata files
load("../Scripts/data/tidy/mobility.RData")
load("../Scripts/data/tidy/daily_aqi_and_pollutants.RData")

# Create mobility dataframe with selected values
mobility_df <- all_selected_counties %>%
  select(date, county, m50, m50_index)

# Create aqi dataframe with selected values
aqi_df <- aqi_df %>%
  filter(date >= "2020/03/01") %>%
  select(date, county, aqi, main_pollutant, ozone, pm25, co, so2, pm10, no2)

# Combine dataframes together and match by date and county
df <- merge(mobility_df, aqi_df, by = c("date", "county"))
```

Next, I seperated the counties by mobility percentage as we did previously but this time with mobility and aqi values included. Finally, I saved it as a the final .Rdata file to use for our analyses.

```{r}
# counties that decreased by <30%
high_mobility <- df[(df$county %in% c("New York", "Arlington", "San Francisco", "Santa Clara", "District of Columbia")), ]

# counties that decreased by 98-100%
low_mobility <- df[(df$county %in% c("Miami-Dade", "Dallas", "Harris", "Maricopa", "San Bernardino")), ]

# save the final dataset to use for analysis
# save(df, low_mobility, high_mobility, file = "data/tidy/tidy_mobility_vs_aqi.RData")

head(df)
head(high_mobility)
head(low_mobility)
```

